---
title: "Worksheet 2"
author: "Reto Zihlmann"
date: "08 October 2019"
output: html_document
---

## Problem 3

I downloaded the frequency.dat file from the [repository](https://github.com/karlrupp/microprocessor-trend-data/tree/master/42yrs).


```{r}
data <- read.table("frequency.dat")
names(data) <- c("year", "freq")
plot(freq ~ year, data = data, ylab = "Freqency [MHz]")
```

A first inspection of the data shows that the data show that the variability frequency increases with the level of frequency and all frequencies are naturally positive. Therefore, a log transformation might be useful to stabilize the variance.

```{r}
plot(log(freq) ~ year, data = data, ylab = "Freqency [MHz]")
```

We try to fit a GAM to the data.

```{r}
library(mgcv)
fit <- gam(log(freq) ~ s(year), data = data)
plot(log(freq) ~ year, data = data, ylab = "Freqency [MHz]")
lines(data$year,predict(fit), col = "red", lwd = 2)
```

GAM is able to follow the trend without overfitting to the single observations.

```{r}
summary(fit)
```

The predictor year is highly significant with `r summary(fit)$edf` estimated degrees of freedom. GAM can also be used to predict future frequencies.

```{r}
plot(log(freq) ~ year, data = data, ylab = "Freqency [MHz]", xlim = c(1970, 2050), ylim = c(0,10))
lines(data$year,predict(fit), col = "red", lwd = 2)
pred_period <- 2018:2050
mypred <- predict.gam(fit, newdata = list(year = pred_period), se.fit = T)
polygon(x = c(pred_period, rev(pred_period)),
        y = c(mypred$fit - 1.96*mypred$se.fit,
              rev(mypred$fit + 1.96*mypred$se.fit)),
        border = NA, col = "lightgray")
lines(pred_period, mypred$fit, col = "gray", lwd = 2)
```


