---
title: "Worksheet 1"
author: "Reto Zihlmann"
date: "24 September 2019"
output: html_document
---

# Problem 1

## (b)

$$
f(x_1, x_2)=10^6x_2^{-6}\exp\left(\frac{-1}{2x_2^2}\left(120 - 6x_1+x_1^2\right)\right)
$$


The equation is similar to the density of the normal distribtution, with $x_2$ representing the variance $\sigma^2$.

The following solution was found with the help of Naim de Viragh.

### (i)

First we define the function as well as the indicater version of the function.

```{r}
myfun <- function(x1, x2 = 4){
  10^6 * x2^-6 * exp(-1/(2*x2^2) * (120 - 6*x1 + x1^2))
}

ind_myfun <- function(x1, x2){
  as.numeric(myfun(x1, x2) >= 2)
}
```

Also we need to set a reasonable range over which we calculate the integral. Outside of this range, the function value is close to zero.

```{r}
dx <- 0.1
x1range <- seq(-10, 15, dx)
x2range <- seq(1, 20, dx)
```

We can display the region where the indicator function is 1.

```{r}
val <- outer(x1range, x2range, FUN = ind_myfun)
image(x = x1range, y = x2range, z = val)
```

Finally we calculate the integral

```{r}
sum(val) * dx * dx
```


### (ii)

Again we display the region and the function value over which we integrate

```{r}
val <- outer(x1range, x2range, FUN = myfun)
val[val < 2] <- 0
image(x = x1range, y = x2range, z = val)
contour(x = x1range, y = x2range, z = val, levels = 2, add = T)
```

We calculate the integral the same way as in (i)

```{r}
sum(val) * dx * dx
```

