---
project: CompSkills    #################################################
title:   Worksheet 6
author:  Reto Zihlmann
date:    2019-12-03 18:15:49
output:  html_document   #############################################
---



## Problem 8

### (a)

I will use a bootstrap approach. The function `stat` samples some of the rows from the data matrix, fits two models (GAM and logistic function) and calculates the $RMSE$ based on the out of sample points.

```{r}
set.seed(1)
library(tidyverse)
library(mgcv)
library(boot)
data <- read.table("frequency.dat")
names(data) <- c("year", "freq")

stat <- function(data, index){
  tryCatch({
    train <- data[index,]
    test <- data[-index,]
    if(!is_empty(test)){
      fit_gam <- gam(log(freq) ~ s(year), data = train)
      fit_ssfpl <- nls(freq ~ SSfpl(year, A, B, xmid, scal), data = train)
      gam_pred <- unname(exp(predict(fit_gam, newdata = test)))
      ssfpl_pred <- predict(fit_ssfpl, newdata = test)
      gam_rmse <- sqrt(mean((test$freq - gam_pred)^2))
      ssfpl_rmse <- sqrt(mean((test$freq - ssfpl_pred)^2))
      rmse <- c(gam_rmse, ssfpl_rmse)
      rmse
    } else {
      stop("no points in test set")
    }
  }, error = function(err) {return(c(NA, NA))})
}
```

Now we run this 100 times.

```{r}
myboot <- boot(data = data, statistic = stat, R = 100)
(rmse_mean <- apply(myboot$t, 2, mean, na.rm = T))
```

We see that the logistic function has a lower $RMSE$, however the fitting process does not always converge. This calculation already takes some time. The function `boot` of package `boot` has a built in option to run the computation in parallel. Let's see if this speeds up our calculation

```{r}
microbenchmark(
  boot_with_parallel = boot(data = data, statistic = stat, R = 10, parallel = c("snow")),
  boot_without_parallel = boot(data = data, statistic = stat, R = 10, parallel = c("no"))
  )
```

The calculation seems not to be faster. However, we used a very small value for R. With larger values this might change. We finally compare the two methodes with a larger number of replications, which takes a long time to compute.

```{r}
myboot <- boot(data = data, statistic = stat, R = 2000, parallel = c("snow"))
(rmse_mean <- apply(myboot$t, 2, mean, na.rm = T))
```



### (b)

The following shell script runs the same calculation as above 4 times with different seeds and saves the $RMSE$ to different `RData` files.

```bash
#!/bin/sh
for seed in 1 4 
do 
   echo $seed   
   Rscript -e "set.seed($seed); source('reto_script.R');save(rmse_mean, file='reto_script_$seed.RData')" &
done
```

Where `reto_script.R` is the following R script

```r
library(tidyverse)
library(mgcv)
library(boot)
data <- read.table("frequency.dat")
names(data) <- c("year", "freq")

stat <- function(data, index){
  tryCatch({
    train <- data[index,]
    test <- data[-index,]
    if(!is_empty(test)){
      fit_gam <- gam(log(freq) ~ s(year), data = train)
      fit_ssfpl <- nls(freq ~ SSfpl(year, A, B, xmid, scal), data = train)
      gam_pred <- unname(exp(predict(fit_gam, newdata = test)))
      ssfpl_pred <- predict(fit_ssfpl, newdata = test)
      gam_rmse <- sqrt(mean((test$freq - gam_pred)^2))
      ssfpl_rmse <- sqrt(mean((test$freq - ssfpl_pred)^2))
      rmse <- c(gam_rmse, ssfpl_rmse)
      rmse
    } else {
      stop("no points in test set")
    }
  }, error = function(err) {return(c(NA, NA))})
}
myboot <- boot(data = data, statistic = stat, R = 2000, parallel = c("snow"))
(rmse_mean <- apply(myboot$t, 2, mean, na.rm = T))
```
